Polynomial Time and Dependent Types

### 2023-10-24 paper/paper.tex
[ paper ] various small fixes

This paper makes the following contributions to the theory and use of
linear dependent type theory and implicit computational complexity:
\begin{enumerate}
\item We formulate two systems that combine linear type theory for
  polytime computation with full dependent types, using Quantitative
  Type Theory. The system is presented in \autoref{sec:qtt}. The
  linear typing discipline required for enforcing polytime is provided
  by QTT, but we also need to carefully add constructs for
  non-iterable datatypes (\autoref{sec:noniter-qtt}) and the two kinds
  of natural number iterator that we consider
  (\autoref{sec:cons-free-qtt} and \autoref{sec:lfpl-qtt}). Porting
  the natural number iterators from the simply typed to the
  dependently typed setting requires careful annotation of the rules
  to ensure that the correct information is available for type
  checking, while also not allowing too much information to be made
  available at runtime that would violate the polytime soundness
  property. A further contribution of this paper is the addition of
  reflection types to QTT, \autoref{sec:qtt-reflection}, which allow
  statements about polytime realisability to be reflected into
  types. % We explain in \autoref{Sec:realisability-future} how this
  % will have applications in QTT beyond the polytime systems presented
  % here.
\item We demonstrate the utility of the combination of polytime and
  dependent types in \autoref{sec:programming-polytime}. Just as in
  the simply typed world, we have an expressive language for writing
  polytime programs. With the additional power of dependent types, we
  can also prove properties of these programs. A simple example is
  proving that a sorting program actually sorts. Using QTT reflection,
  we can go further and represent the class of polytime problems, with
  polytime reductions between them, as dependent pairs
  (\autoref{sec:polytime-problems}). Our final examples use dependent
  types to give monadic presentations of the complexity classes of
  Non-deterministic Polynomial (NP) time and Probabilistic Polynomial
  (PP) time. Since these classes rely on specific correctness
  criteria, it is not possible to capture them in a simply typed
  system for polytime.
\item We prove the polytime soundness of our systems via a
  realisability argument in \autoref{sec:soundness}. Our construction
  is an extension of the amortised complexity realisability
  constructions of \citet{dallago11realisability}. We extend their
  work to our dependently typed setting, and also give a realisability
  interpretation of datatypes directly, instead of via second-order
  impredicative encodings. The technical content of this section has
  been formalised in the Agda proof assistant
  \cite{norell2008dependently}, and is included in the supplementary
  material.
\end{enumerate}

@@

Integrating linear and dependent types is not straightforward due to
the conflict between the linear typing discipline regarding presence
of a variable as only bestowing the right to use it once, and the
dependent typing regime that uses variables both in types (for
specification purposes) and in terms (for computational purposes),
syntactically yield multiple uses of the same variable.

@@

We use an instantiation of QTT with the natural number semiring, with
the usual semiring structure of addition and multiplication. In a mild
generalisation of the original presentation of QTT, we also allow
sub-usaging via the \emph{reverse} ordering on the naturals. That is,
if a variable is marked as usage $n$ and $m \geq n$, then we can also
regard it as usage $m$. This makes the system more like affine linear
logic, since $m \leq 0$ for all $m$, matching the system in
\autoref{sec:linear-intro}. We do \emph{not} have an unrestricted
usage $\omega$, since this would allow the possibility of unrestricted
duplication, and hence violate our polytime soundness properties.

@@

$\Sigma$-types are a little more involved, and demonstrate the
flexibility in QTT in allowing additional power in the $\sigma = 0$
fragment, where we do not need to care about polytime
realisability. Formation and introduction are given by the rules:
\begin{mathpar}
  \inferrule*
  {0\Gamma \vdash S \\ 0\Gamma, x \stackrel0: S \vdash T}
  {0\Gamma \vdash (x \stackrel\pi: S) \otimes T}

  \inferrule*
  {\Gamma_1 \vdash M \stackrel{\sigma'}: S \\
    \Gamma_2 \vdash N \stackrel\sigma: T[M/x] \\\\
    0\Gamma_1 = 0\Gamma_2 \\
    \sigma' = 0 \Leftrightarrow (\pi = 0 \lor \sigma = 0)}
  {\pi\Gamma_1 + \Gamma_2 \vdash (M, N) \stackrel\sigma: (x \stackrel\pi: S) \otimes T}
\end{mathpar}
As with $\Pi$-typed, the first component of a $\Sigma$ type is
annotated with a usage for how many times it can be used, and this is
respected by the introduction rule. Elimination of $\Sigma$-types
depends on whether we are in the $\sigma = 0$ fragment or not. In the
$\sigma = 0$ fragment, we are free to disregard usage restrictions,
and use projections as normal:
\begin{mathpar}
  \inferrule*
  {\Gamma \vdash M \stackrel0: (x \stackrel\pi: S) \otimes T}
  {\Gamma \vdash \mathrm{fst}(M) \stackrel0: S}

@@

  \inferrule*
  {\Gamma_1 \vdash M \stackrel\sigma: T \\
    \Gamma_2 \vdash N \stackrel\sigma: \ListTy(T) \\
    0\Gamma_1 = 0\Gamma_2}
  {\Gamma_1 + \Gamma_2 \vdash \cCons(M,N) \stackrel\sigma: \ListTy(T)}
\end{mathpar}
Lists do have the potential for iteration by their recursive nature,
but in order to ensure the complexity guarantees we only permit
matching without recursion in the full calculus. Here is the rule for
dependently typed case analysis on lists, which also obeys the usual
$\beta$-equalities for case analysis, analogous to the ones for
Booleans:
\begin{mathpar}
  \inferrule*
  {0\Gamma_1, x \stackrel0: \ListTy(T) \vdash P~\istype \\
    \Gamma_1 \vdash M \stackrel\sigma: \ListTy(T) \\
    \Gamma_2 \vdash N_1 \stackrel\sigma: P[\cNil/x] \\
    \Gamma_2, h \stackrel\sigma: T, t \stackrel\sigma: \ListTy(T) \vdash N_2 \stackrel\sigma: P[\cCons(h,t)/x] \\
    0\Gamma_1 = 0\Gamma_2}
  {\Gamma_1 + \Gamma_2 \vdash \Match_{x.P}\,M\,\{\,\cNil \mapsto N_1; \cCons(h,t) \mapsto N_2\,\} \stackrel\sigma: P[M/x]}
\end{mathpar}
In the $\sigma = 0$ fragment, however, we are free to iterate on lists
because computations in this fragment are only meant for type-level
computation, not for the program itself. Put another way, the type
checker may perform arbitary recursion on lists to type check the
program, but the program itself may not do so without correctly
accounting its costs as described in the following sections. The
$\sigma = 0$ recursor for lists has the following typing rule, which
is the standard dependent eliminator for lists, with everything
annotated as $0$ usage.
\begin{mathpar}
  \inferrule*
  {0\Gamma, x \stackrel0: \ListTy(T) \vdash P~\istype \\
    0\Gamma \vdash M \stackrel0: \ListTy(T) \\
    0\Gamma \vdash N_1 \stackrel0: P[\cNil/x] \\
    0\Gamma, h \stackrel0: T, t \stackrel0: \ListTy(T), p \stackrel0: P[t/x] \vdash N_2 \stackrel0: P[\cCons(h,t)/x]}
  {0\Gamma \vdash \mathrm{recList}_{x.P}\,M\,\{\,\cNil \mapsto N_1; \cCons(h,t;p) \mapsto N_2\,\} \stackrel0: P[M/x]}
\end{mathpar}
This eliminator also obeys the usual $\beta$-equality laws for a list
eliminator, using the resource freedom of the $\sigma = 0$ fragment to
duplicate the $N_2$ term in the $\cCons$ case.

@@

The cons-free system cannot allow the programmer to construct natural
numbers in the $\sigma = 1$ fragment, as this would violate the
complexity guarantees. However, we can use the flexibility of QTT to
allow free construction of naturals in the $\sigma = 0$ fragment,
which allows us to use natural numbers freely in types. Therefore, we
have the following introduction rules, only usable in the $\sigma = 0$
fragment:
\begin{mathpar}
  \inferrule*
  {\Gamma \vdash}
  {\Gamma \vdash \conZero \stackrel0: \tyNat}

@@

As explained in \autoref{sec:lfpl-intro}, the LFPL system differs from
the Cons-free system in that it is possible to construct natural
numbers (and other iterable datatypes), provided one has the necessary
diamonds to pay for the construction. As with the natural number type
in the Cons-free system, it ought not be possible to construct
diamonds in the $\sigma = 1$ fragment, as this would amount to the
free distribution of diamonds to all which would lead to a collapse in
the complexity guarantees of the system. It is possible construct
diamonds in the $\sigma = 0$, though:
\begin{mathpar}
  \inferrule*
  {\Gamma \vdash}
  {0\Gamma \vdash \Diamond~\istype}

  \inferrule*
  {\Gamma \vdash}
  {0\Gamma \vdash * \stackrel0: \Diamond}

  \inferrule*
  {\Gamma \vdash M \stackrel0: \Diamond}
  {\Gamma \vdash M \equiv * \stackrel0: \Diamond}
\end{mathpar}
The $\Diamond$ type also supports an $\eta$-rule in the $\sigma = 0$
fragment, indicating that, in this fragment, it acts the same as a
unit type. This allows us to use freely use diamonds in types, and to
not have to care about the identity of particular diamonds, since by
this rule all diamonds are definitionally
equal\footnote{\emph{Fungible}, if one wishes to use a monetary
  metaphor.}.
% Link to Selinger et al: the $\eta$ rule for $\Diamond$s means that are
% essentially depending on the ``shape'' of anything that depends on
% diamonds.

@@

We have only defined an iterable natural number datatype for both of
our systems above. We could extend both systems to include further
iterable inductive types, although in the \emph{Cons-free} system this
is not particularly useful due to the prohibition of on
construction. However, sticking with just the natural numbers, we can
use the power of dependent types with a universe to create further
datatypes whose size is measured by some iterable natural
number. Iteration on the size yields iteration over the full
datastructure. For example, in the LFPL system, we can define a type
of iterable lists by pairing a size with a type of elements defined by
recursion on the size:
\begin{displaymath}
  \mathrm{IList}\,A = (n \stackrel1: \tyNat) \otimes \left(\mathrm{rec}_{x.\mathsf{U}}\,n\,\{ \conZero(d) \mapsto I; \conSucc(d,n;p) \mapsto A \otimes p \}\right)
\end{displaymath}
The $\cNil$ and $\cCons$ constructors can now be defined in terms of
$\conZero$ and $\conSucc$, provided the caller supplies sufficient
$\Diamond$s. These definitions live in the $\sigma = 1$ fragment, so
we annotate them appropriately:
\begin{mathpar}
  \begin{array}{l}
    \mathrm{nil} \stackrel1: \Diamond \to \mathrm{IList}\,A \\
    \mathrm{nil}\,d = (\conZero(d), *)
  \end{array}

  \begin{array}{l}
    \mathrm{cons} \stackrel1: \Diamond \to A \to \mathrm{IList}\,A \to \mathrm{IList}\,A \\
    \mathrm{cons}\,d\,x\,\mathit{xs} = \mathrm{let}\,(n,\mathit{elems}) = \mathit{xs}\,\textrm{in}\,(\conSucc(d,n),(x,\mathit{elems}))
  \end{array}
\end{mathpar}
Using the LFPL iterator it is also possible to construct a dependently
typed iterator for $\mathrm{IList}\,A$ values. Unfortunately, the
current types of the LFPL system are not sufficient to type this as a
function, as we have no way of stating that the successor case must be
arbitrarily duplicable. Lifting this restrict by means of some
modality is future work. The typing rule for the derived list iterator
is:
\begin{mathpar}
  \mprset{flushleft}
  \inferrule*
  {0\Gamma \vdash A~\istype \\
    0\Gamma, x \stackrel0: \mathrm{IList}\,A \vdash P~\istype \\\\
    \Gamma \vdash M \stackrel\sigma: \mathrm{IList}\,A \\\\
    0\Gamma, d \stackrel1: \Diamond \vdash N_1 \stackrel\sigma: P[\mathrm{nil}(*)/x] \\\\
    0\Gamma, d \stackrel1: \Diamond, x \stackrel\sigma: A, \mathrm{xs} \stackrel0: \mathrm{IList}\,A, p \stackrel\sigma: P[\mathit{xs}/x] \vdash N_2 \stackrel\sigma : P[\mathrm{cons}(*,x,\mathit{xs})/x]}
  {\Gamma \vdash \tmRec_{x.P}\,M\,\{\mathrm{nil}(d) \mapsto N_1; \mathrm{cons}(d,x,\mathit{xs};p) \mapsto N_2\} \stackrel\sigma: P[M/x]}
\end{mathpar}
Note that, in the cons case, we have access to the result of iterating
over the tail of the list ($p$), but not to the actual tail of the
list ($\mathit{xs}$).

@@

We can use the reflection type former defined in
\autoref{sec:qtt-reflection} to define a predicate on decision
problems that establishes whether or not they are polytime decision
problems. Specifically, we can state that there is a polytime
realisable boolean-value predicate that reports true exactly when the
given element of $a$ is in the predicate:
\begin{displaymath}
  \mathrm{PTIME}(A,P) =
    (f \stackrel1: \mathbf{R}(A \to \BoolTy)) \otimes (a \stackrel1: A) \to (\mathbf{R}^{-1}(f)\, a = \cTrue) \Leftrightarrow P\,a
\end{displaymath}
Thus, $\mathrm{PTIME}(A,P)$ is a logical proposition stating that the
decision problem $(A,P)$ is deciable in polytime. We make two notes
about this definition: (i) this definition is intrinsic, in the sense
that, whichever of the polytime systems is chosen, proving that a
decision problem is solvable in polytime is a matter of programming,
without having to reason directly about machine models and step
counting; and (ii) we have defined problems to have arbitrary types
$A$ as domains, rather than bitstrings, and so the notion of size
attached to an input is intrinsic to the type $A$ chosen.

@@

\subsection{Polytime-based Complexity Classes}

@@

The type $\mathrm{ND}$ supports a monad interface via the usual free
monad construction, as well as an effect
$\mathrm{flip} \stackrel1: \mathrm{ND}\,\BoolTy$ providing access to a
bit of non-deterministic information. Thus a program of type
$A \to \mathrm{ND}\,B$ in the $\sigma = 1$ fragment will be a polytime
program with access to an oracle. In the $\sigma = 0$ fragment, we can
write a function that resolves non-determinism using a list of
booleans. This function returns $\mathsf{nothing}$ if the list of
booleans is insufficient to resolve all the $\mathsf{choice}$s:
\begin{displaymath}
  \mathrm{runWithOracle} \stackrel0: \mathrm{ND}\,A \to \ListTy(\BoolTy) \to \mathrm{Maybe}\,A
\end{displaymath}
With these definitions, we can define Non-deterministic Polynomial
time as a predicate on problems:
\begin{displaymath}
  \mathrm{NP}(A,P) =
  \begin{array}[t]{@{}l}
    (f \stackrel1: \mathbf{R}(A \to \mathrm{ND}(\BoolTy)))\, \otimes \\
    \quad (a \stackrel1: A) \to \left((\mathit{bs} \stackrel1: \ListTy(\BoolTy)) \otimes (\mathrm{runWithOracle}\,(\mathbf{R}^{-1}(f)\,a)\,\mathit{bs} = \mathsf{just}\,\cTrue)\right) \Leftrightarrow P\,a
  \end{array}
\end{displaymath}
Thus, a problem is in NP if there is a non-deterministic
boolean-valued polynomial time function that has a path to returning
$\cTrue$ exactly when the input satisfies the predicate. Moreover, it is a
quick matter of programming to see that problems in NP are closed
under the type of polytime reductions given above.

@@

\paragraph{Agda Formalisation} The machine model is defined in the
Agda module \AgdaModule{MachineModel}. We use an intrinsicaly
well-scoped syntax, which ensures that all variable look up operations
are always well defined.

@@

\begin{definition}\label{def:indexed-linear-preorder}
  A \emph{$\mathbb{N}$-linear preorder}\footnote{We specialise to the semiring $\mathbb{N}$ here, but the same definition works for any suitable semiring $\mathcal{R}$.} is a preordered set
  $(L, \leq)$ with
  \begin{enumerate}
  \item a commutative monoid $(I, -\otimes-)$ that is monotone
    w.r.t. the order;
  \item is closed: there is an operation
    $\multimap : L \times L \to L$ such that $x \otimes y \leq z$ iff
    $x \leq y \multimap z$; and
  \item has a function $\oc : \mathbb{N} \to L \to L$, to interpret
    resource requirement adjustments, satisfying:
    \begin{enumerate}
    \item $\oc_0 X \simeq I$, for discarding;
    \item $\oc_{m+n} X \leq (\oc_m X) \otimes (\oc_n X)$, for duplication;
    \item $\oc_m\oc_n X \leq \oc_{mn} X$ for nesting;
    \item $\oc_1 X \leq X$ for extraction / dereliction;
    \item $(\oc_n X) \otimes (\oc_n Y) \leq \oc_n(X \otimes Y)$, for distribution; and
    \item $n \leq m$ implies $\oc_n X \leq \oc_m X$, for usage weakening.
    \end{enumerate}
  \end{enumerate}
  The collection of all linear preorders and functions that preserve the
  order and the operations forms a category $\LinPreorder$.

@@

\paragraph{Booleans} Fix $\mathbb{B} = \{ \mathit{tt}, \mathit{ff} \}$
as our set of boolean elements. We define an element of
$L(\mathbb{B})$ to represent boolean values:
\begin{displaymath}
  \mathrm{Bool} = \{ (\mathit{tt}, \alpha, \synTrue) \mid \alpha \in M \} \cup \{ (\mathit{ff}, \alpha, \synFalse) \mid \alpha \in M \}
\end{displaymath}
Thus, the boolean $\mathit{tt}$ is represented by the value $\synTrue$
and $\mathit{ff}$ is represented by $\synFalse$. In both case, we
allow arbitrary potential $\alpha$ to be attached.

>>>>

This paper makes the following contributions to the theory and use of
linear dependent type theory and implicit computational complexity:
\begin{enumerate}
\item We formulate two systems that combine linear type theory for
  polytime computation with full dependent types, using Quantitative
  Type Theory. The systems are presented in \autoref{sec:qtt}. The
  linear typing discipline required for enforcing polytime is provided
  by QTT, but we also need to carefully add constructs for
  non-iterable datatypes (\autoref{sec:noniter-qtt}) and the two kinds
  of natural number iterator that we consider
  (\autoref{sec:cons-free-qtt} and \autoref{sec:lfpl-qtt}). Porting
  the natural number iterators from the simply typed to the
  dependently typed setting requires careful annotation of the rules
  to ensure that the correct information is available for type
  checking, while also not allowing too much information to be made
  available at runtime that would violate the polytime soundness
  property. A further contribution of this paper is the addition of
  reflection types to QTT, \autoref{sec:qtt-reflection}, which allow
  statements about polytime realisability to be reflected into
  types. % We explain in \autoref{Sec:realisability-future} how this
  % will have applications in QTT beyond the polytime systems presented
  % here.
\item We demonstrate the utility of the combination of polytime and
  dependent types in \autoref{sec:programming-polytime}. Just as in
  the simply typed world, we have an expressive language for writing
  polytime programs. With the additional power of dependent types, we
  can also prove properties of these programs. A simple example is
  proving that a sorting program actually sorts. Using QTT reflection,
  we can go further and represent the class of polytime problems, with
  polytime reductions between them, as dependent pairs
  (\autoref{sec:polytime-problems}). Our final examples use dependent
  types to give monadic presentations of the complexity classes of
  Non-deterministic Polynomial (NP) time and Probabilistic Polynomial
  (PP) time. Since these classes rely on specific correctness
  criteria, it is not possible to capture them in a simply typed
  system for polytime.
\item We prove the polytime soundness of our systems via a
  realisability argument in \autoref{sec:soundness}. Our construction
  is an extension of the amortised complexity realisability
  constructions of \citet{dallago11realisability}. We extend their
  work to our dependently typed setting, and also give a realisability
  interpretation of datatypes directly, instead of via second-order
  impredicative encodings. The technical content of this section has
  been formalised in the Agda proof assistant
  \cite{norell2008dependently}, and is included in the supplementary
  material.
\end{enumerate}

@@

Integrating linear and dependent types is not straightforward due to
the conflict between the linear typing discipline regarding presence
of a variable as only bestowing the right to use it once, and the
dependent typing regime that uses variables both in types (for
specification purposes) and in terms (for computational purposes),
syntactically yielding multiple uses of the same variable.

@@

We use an instantiation of QTT with the natural number semiring, with
the usual semiring structure of addition and multiplication. In a mild
generalisation of the original presentation of QTT, we also allow
sub-usaging via the \emph{reverse} ordering on the naturals. That is,
if a variable is marked as usage $n$ and $m \geq n$, then we can also
regard it as usage $m$. This makes the system more like affine linear
logic, since $m \leq 0$\footnote{Reverse ordering!} for all $m$,
matching the system in \autoref{sec:linear-intro}. We do \emph{not}
have an unrestricted usage $\omega$, since this would allow the
possibility of unrestricted duplication, and hence violate our
polytime soundness properties.

@@

$\Sigma$-types are a little more involved, and demonstrate the
flexibility in QTT in allowing additional power in the $\sigma = 0$
fragment, where we do not need to care about polytime
realisability. Formation and introduction are given by the rules:
\begin{mathpar}
  \inferrule*
  {0\Gamma \vdash S~\istype \\ 0\Gamma, x \stackrel0: S \vdash T~\istype}
  {0\Gamma \vdash (x \stackrel\pi: S) \otimes T~\istype}

  \inferrule*
  {\Gamma_1 \vdash M \stackrel{\sigma'}: S \\
    \Gamma_2 \vdash N \stackrel\sigma: T[M/x] \\\\
    0\Gamma_1 = 0\Gamma_2 \\
    \sigma' = 0 \Leftrightarrow (\pi = 0 \lor \sigma = 0)}
  {\pi\Gamma_1 + \Gamma_2 \vdash (M, N) \stackrel\sigma: (x \stackrel\pi: S) \otimes T}
\end{mathpar}
As with $\Pi$-types, the first component of a $\Sigma$-type is
annotated with a usage for how many times it can be used, and this is
respected by the introduction rule. Elimination of $\Sigma$-types
depends on whether we are in the $\sigma = 0$ fragment or not. In the
$\sigma = 0$ fragment, we are free to disregard usage restrictions,
and use projections as normal:
\begin{mathpar}
  \inferrule*
  {\Gamma \vdash M \stackrel0: (x \stackrel\pi: S) \otimes T}
  {\Gamma \vdash \mathrm{fst}(M) \stackrel0: S}

@@

  \inferrule*
  {\Gamma_1 \vdash M \stackrel\sigma: T \\
    \Gamma_2 \vdash N \stackrel\sigma: \ListTy(T) \\
    0\Gamma_1 = 0\Gamma_2}
  {\Gamma_1 + \Gamma_2 \vdash \cCons(M,N) \stackrel\sigma: \ListTy(T)}
\end{mathpar}
Lists do have the potential for iteration by their recursive nature,
but in order to ensure the complexity guarantees we only permit
matching without recursion in the full calculus. Here is the rule for
dependently typed case analysis on lists, which also obeys the usual
$\beta$-equalities for case analysis, analogous to the ones for
Booleans:
\begin{mathpar}
  \inferrule*
  {0\Gamma_1, x \stackrel0: \ListTy(T) \vdash P~\istype \\
    \Gamma_1 \vdash M \stackrel\sigma: \ListTy(T) \\
    \Gamma_2 \vdash N_1 \stackrel\sigma: P[\cNil/x] \\
    \Gamma_2, h \stackrel\sigma: T, t \stackrel\sigma: \ListTy(T) \vdash N_2 \stackrel\sigma: P[\cCons(h,t)/x] \\
    0\Gamma_1 = 0\Gamma_2}
  {\Gamma_1 + \Gamma_2 \vdash \Match_{x.P}\,M\,\{\,\cNil \mapsto N_1; \cCons(h,t) \mapsto N_2\,\} \stackrel\sigma: P[M/x]}
\end{mathpar}
In the $\sigma = 0$ fragment, however, we are free to iterate on lists
because computations in this fragment are only meant for type-level
computation, not for the program itself. Put another way, the type
checker may perform arbitary recursion on lists to type check the
program, but the program itself may not do so without correctly
accounting its costs as described in the following sections. The
$\sigma = 0$ fragment recursor for lists has the following typing
rule, which is the standard dependent eliminator for lists, with
everything annotated as $0$ usage.
\begin{mathpar}
  \inferrule*
  {0\Gamma, x \stackrel0: \ListTy(T) \vdash P~\istype \\
    0\Gamma \vdash M \stackrel0: \ListTy(T) \\
    0\Gamma \vdash N_1 \stackrel0: P[\cNil/x] \\
    0\Gamma, h \stackrel0: T, t \stackrel0: \ListTy(T), p \stackrel0: P[t/x] \vdash N_2 \stackrel0: P[\cCons(h,t)/x]}
  {0\Gamma \vdash \mathrm{recList}_{x.P}\,M\,\{\,\cNil \mapsto N_1; \cCons(h,t;p) \mapsto N_2\,\} \stackrel0: P[M/x]}
\end{mathpar}
This eliminator also obeys the usual $\beta$-equality laws for a list
eliminator, using the resource freedom of the $\sigma = 0$ fragment to
duplicate the $N_2$ term in the $\cCons$ case.

@@

The cons-free system cannot allow the programmer to construct natural
numbers in the $\sigma = 1$ fragment, as this would violate the
complexity guarantees. However, we can use the flexibility of QTT to
allow free construction of naturals in the $\sigma = 0$ fragment,
which allows us to use natural numbers freely in types. Therefore, we
have the following introduction rules, only usable in the $\sigma = 0$
fragment:
\begin{mathpar}
  \inferrule*
  {\Gamma~\isctxt}
  {\Gamma \vdash \conZero \stackrel0: \tyNat}

@@

As explained in \autoref{sec:lfpl-intro}, the LFPL system differs from
the Cons-free system in that it is possible to construct natural
numbers (and other iterable datatypes), provided one has the necessary
diamonds to pay for the construction. As with the natural number type
in the Cons-free system, it ought not be possible to construct
diamonds in the $\sigma = 1$ fragment, as this would amount to the
free distribution of diamonds to all which would lead to a collapse in
the complexity guarantees of the system. It is possible construct
diamonds in the $\sigma = 0$, though:
\begin{mathpar}
  \inferrule*
  {\Gamma~\isctxt}
  {0\Gamma \vdash \Diamond~\istype}

  \inferrule*
  {\Gamma~\isctxt}
  {0\Gamma \vdash * \stackrel0: \Diamond}

  \inferrule*
  {\Gamma \vdash M \stackrel0: \Diamond}
  {\Gamma \vdash M \equiv * \stackrel0: \Diamond}
\end{mathpar}
The $\Diamond$ type also supports an $\eta$-rule in the $\sigma = 0$
fragment, indicating that, in this fragment, it acts the same as a
unit type. This allows us to freely use diamonds in types, and to
not have to care about the identity of particular diamonds, since by
this rule all diamonds are definitionally
equal\footnote{\emph{Fungible}, if one wishes to use a monetary
  metaphor.}.
% Link to Selinger et al: the $\eta$ rule for $\Diamond$s means that are
% essentially depending on the ``shape'' of anything that depends on
% diamonds.

@@

We have only defined an iterable natural number datatype for both of
our systems above. We could extend both systems to include further
iterable inductive types, although in the \emph{Cons-free} system this
is not particularly useful due to the prohibition of
construction. However, sticking with just the natural numbers, we can
use the power of dependent types with a universe to create further
datatypes whose size is measured by some iterable natural
number. Iteration on the size yields iteration over the full
datastructure. For example, in the LFPL system, we can define a type
of iterable lists by pairing a size with a type of elements defined by
recursion on the size:
\begin{displaymath}
  \mathrm{IList}\,A = (n \stackrel1: \tyNat) \otimes \left(\mathrm{rec}_{x.\mathsf{U}}\,n\,\{ \conZero(d) \mapsto I; \conSucc(d,n;p) \mapsto A \otimes p \}\right)
\end{displaymath}
The $\cNil$ and $\cCons$ constructors can now be defined in terms of
$\conZero$ and $\conSucc$, provided the caller supplies sufficient
$\Diamond$s. These definitions live in the $\sigma = 1$ fragment, so
we annotate them appropriately:
\begin{mathpar}
  \begin{array}{l}
    \mathrm{nil} \stackrel1: \Diamond \to \mathrm{IList}\,A \\
    \mathrm{nil}\,d = (\conZero(d), *)
  \end{array}

  \begin{array}{l}
    \mathrm{cons} \stackrel1: \Diamond \to A \to \mathrm{IList}\,A \to \mathrm{IList}\,A \\
    \mathrm{cons}\,d\,x\,\mathit{xs} = \mathrm{let}\,(n,\mathit{elems}) = \mathit{xs}\,\textrm{in}\,(\conSucc(d,n),(x,\mathit{elems}))
  \end{array}
\end{mathpar}
Using the LFPL iterator it is also possible to construct a dependently
typed iterator for $\mathrm{IList}\,A$ values. Unfortunately, the
current types of the LFPL system are not sufficient to type this as a
function, as we have no way of stating that the successor case must be
arbitrarily duplicable. Lifting this restriction by means of some
modality is future work. The typing rule for the derived list iterator
is:
\begin{mathpar}
  \mprset{flushleft}
  \inferrule*
  {0\Gamma \vdash A~\istype \\
    0\Gamma, x \stackrel0: \mathrm{IList}\,A \vdash P~\istype \\\\
    \Gamma \vdash M \stackrel\sigma: \mathrm{IList}\,A \\\\
    0\Gamma, d \stackrel1: \Diamond \vdash N_1 \stackrel\sigma: P[\mathrm{nil}(*)/x] \\\\
    0\Gamma, d \stackrel1: \Diamond, x \stackrel\sigma: A, \mathrm{xs} \stackrel0: \mathrm{IList}\,A, p \stackrel\sigma: P[\mathit{xs}/x] \vdash N_2 \stackrel\sigma : P[\mathrm{cons}(*,x,\mathit{xs})/x]}
  {\Gamma \vdash \tmRec_{x.P}\,M\,\{\mathrm{nil}(d) \mapsto N_1; \mathrm{cons}(d,x,\mathit{xs};p) \mapsto N_2\} \stackrel\sigma: P[M/x]}
\end{mathpar}
Note that, in the cons case, we have access to the result of iterating
over the tail of the list ($p$), but not to the actual tail of the
list ($\mathit{xs}$).

@@

We can use the reflection type former defined in
\autoref{sec:qtt-reflection} to define a predicate on decision
problems that establishes whether or not they are polytime decision
problems. Specifically, we can state that there is a polytime
realisable boolean-value predicate that reports true exactly when the
given element of $a$ is in the predicate:
\begin{displaymath}
  \mathrm{PTIME}(A,P) =
    (f \stackrel1: \mathbf{R}(A \to \BoolTy)) \otimes (a \stackrel1: A) \to (\mathbf{R}^{-1}(f)\, a = \cTrue) \Leftrightarrow P\,a
\end{displaymath}
Thus, $\mathrm{PTIME}(A,P)$ is a logical proposition stating that the
decision problem $(A,P)$ is deciable in polytime. We make three notes
about this definition: (i) \emph{proofs} of $\mathrm{PTIME}(A,P)$, are
carried out in the $\sigma = 0$ fragment, where we have the full power
of Type Theory to aid us; (ii) this definition is intrinsic, in the
sense that, whichever of the polytime systems is chosen, proving that
a decision problem is solvable in polytime is a matter of programming,
without having to reason directly about machine models and step
counting; and (iii) we have defined problems to have arbitrary types
$A$ as domains, rather than bitstrings, and so the notion of size
attached to an input is intrinsic to the type $A$ chosen.

@@

\subsection{Polytime-based Complexity Classes}
\label{sec:classes}

@@

The type $\mathrm{ND}$ supports a monad interface via the usual free
monad construction, as well as an effect
$\mathrm{flip} \stackrel1: \mathrm{ND}\,\BoolTy$ providing access to a
bit of non-deterministic information. Thus a program of type
$A \to \mathrm{ND}\,B$ in the $\sigma = 1$ fragment will be a polytime
program with access to an oracle. In the $\sigma = 0$ fragment, we can
write a function that resolves non-determinism using a list of
booleans. This function returns $\mathsf{nothing}$ if the list of
booleans is insufficient to resolve all the $\mathsf{choice}$s:
\begin{displaymath}
  \mathrm{runWithOracle} \stackrel0: \mathrm{ND}\,A \to \ListTy(\BoolTy) \to \mathrm{Maybe}\,A
\end{displaymath}
With these definitions, we can define Non-deterministic Polynomial
time as a predicate on problems:
\begin{displaymath}
  \mathrm{NP}(A,P) =
  \begin{array}[t]{@{}l}
    (f \stackrel1: \mathbf{R}(A \to \mathrm{ND}(\BoolTy)))\, \otimes \\
    \quad \left((a \stackrel1: A) \to \left((\mathit{bs} \stackrel1: \ListTy(\BoolTy)) \otimes (\mathrm{runWithOracle}\,(\mathbf{R}^{-1}(f)\,a)\,\mathit{bs} = \mathsf{just}\,\cTrue)\right) \Leftrightarrow P\,a\right)
  \end{array}
\end{displaymath}
Thus, a problem is in NP if there is a non-deterministic
boolean-valued polynomial time function that has a path to returning
$\cTrue$ exactly when the input satisfies the predicate. Moreover, it is a
quick matter of programming to see that problems in NP are closed
under the type of polytime reductions given above.

@@

\paragraph{Agda Formalisation} The machine model is defined in the
Agda module \AgdaModule{MachineModel}. We use an intrinsically
well-scoped syntax, which ensures that all variable look up operations
are always well defined.

@@

\begin{definition}\label{def:indexed-linear-preorder}
  A \emph{$\mathbb{N}$-linear preorder}\footnote{We specialise to the semiring $\mathbb{N}$ here, but the same definition works for any suitable semiring $\mathcal{R}$.} is a preordered set
  $(L, \leq)$:
  \begin{enumerate}
  \item a commutative monoid $(I, -\otimes-)$ that is monotone
    w.r.t. the order;
  \item is closed: there is an operation
    $\multimap : L \times L \to L$ such that $x \otimes y \leq z$ iff
    $x \leq y \multimap z$; and
  \item has a function $\oc : \mathbb{N} \to L \to L$, to interpret
    resource requirement adjustments, satisfying:
    \begin{enumerate}
    \item $\oc_0 X \simeq I$, for discarding;
    \item $\oc_{m+n} X \leq (\oc_m X) \otimes (\oc_n X)$, for duplication;
    \item $\oc_m\oc_n X \leq \oc_{mn} X$ for nesting;
    \item $\oc_1 X \leq X$ for extraction / dereliction;
    \item $(\oc_n X) \otimes (\oc_n Y) \leq \oc_n(X \otimes Y)$, for distribution; and
    \item $n \leq m$ implies $\oc_n X \leq \oc_m X$, for usage weakening.
    \end{enumerate}
  \end{enumerate}
  The collection of all linear preorders and functions that preserve the
  order and the operations forms a category $\LinPreorder$.

@@

\paragraph{Booleans} Fix $\mathbb{B} = \{ \mathit{tt}, \mathit{ff} \}$
as our set of boolean elements. We define an element of
$L(\mathbb{B})$ to represent boolean values:
\begin{displaymath}
  \mathrm{Bool} = \{ (\mathit{tt}, \alpha, \synTrue) \mid \alpha \in M \} \cup \{ (\mathit{ff}, \alpha, \synFalse) \mid \alpha \in M \}
\end{displaymath}
Thus, the boolean $\mathit{tt}$ is represented by the value $\synTrue$
and $\mathit{ff}$ is represented by $\synFalse$. In both cases, we
allow arbitrary potential $\alpha$ to be attached.
